```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#libraries
library(dplyr)
library(pROC)
library(ggplot2)
library(plotly)


#open the file
cbb=read.csv("cbb.csv",header=TRUE,sep=",")


#group by conference and year to calculate conference average
conference <- cbb %>%
  group_by(YEAR, CONF) %>%
  mutate(avgwin = mean(W, na.rm=TRUE), avgefg= mean(EFG_O, na.rm=TRUE), avg2p= mean(X2P_O, na.rm=TRUE), avg3p= mean(X3P_O, na.rm=TRUE), avgftr = mean(FTR, na.rm=TRUE), avgorb = mean(ORB, na.rm=TRUE), avgefgd = mean(EFG_D, na.rm=TRUE),avgtor = mean(TOR, na.rm=TRUE), avgtord = mean(TORD, na.rm=TRUE), avgdrb = mean(DRB, na.rm=TRUE), avgftrd = mean(FTRD, na.rm=TRUE), avg2pd = mean(X2P_D, na.rm=TRUE), avg3pd = mean(X3P_D, na.rm=TRUE), avgadj = mean(ADJ_T, na.rm=TRUE)) %>%
  ungroup() 

#calculate ratio of team performance to conference performance
conference <- conference %>% mutate(winrate = (W/avgwin), efgrate= (EFG_O/avgefg), rate2p= (X2P_O/avg2p), rate3p= (X3P_O/avg3p), ftrrate = (FTR/avgftr), orbrate = (ORB/avgorb), efgdrate = (EFG_D/avgefgd), torrate = (TOR/avgtor), tordrate = (TORD/avgtord), drbrate = (DRB/avgdrb), ftrdrate = (FTRD/avgftrd), d2prate = (X2P_D/avg2pd), d3prate = (X3P_D/avg3pd), adjrate = (ADJ_T/avgadj))

#filter outliers + create binary variable for final 4
conferenceclean <- conference %>% filter(!CONF %in% c('Ind', 'ind', 'GWC', 'Sum'))
conferenceclean$finals=ifelse(conferenceclean$POSTSEASON %in% c("Champions","2nd","F4"),1,0)

#Grouping for Top 10 Charts
TopWins <- conference %>%
  group_by(CONF) %>%
  summarize(TotalWins = sum(W))
TopWins <- TopWins %>% 
  arrange(desc(TotalWins)) %>%
  slice(1:10)

TeamWins <- conference %>%
  group_by(TEAM) %>%
  summarize(TotalWins = sum(W), Conference = first(CONF))
TeamWins <- TeamWins %>% 
  arrange(desc(TotalWins)) %>%
  slice(1:10)

```

## Data Interpretation

Put visuals that explore the data here

best team

best overall performance

```{r}
##M

##### Librarys Used
#library(readxl)
#library(ggplot2)
#library(ggcorrplot)
#library(dplyr)
#library(leaps)
#library(forecast)
#library(pROC)
#library(tidyverse)
#library(factoextra)

## Visuals/test/validations

## VISUALIZATIONS
data = cbb

data$WINSEAS <- ifelse(data$W/data$G > .5,1,0)

data$MADETOURN <- ifelse(data$POSTSEASON == 'NA',0,1)

average <- mean(data$W)

data$aboveavg <- ifelse(data$W > average,1,0)

### 
######## TOP THREEE MOST WINNING CONFERENCES EACH YEAR

mostwinningconf <- data%>%group_by(YEAR,CONF)%>%
  summarize(total = sum(W))%>%
  arrange(YEAR,-total)%>%
  slice(1:3)

mostwinningconf

ggplot(data=mostwinningconf, aes(x= YEAR, y=total, fill=CONF)) +
  geom_bar(stat="identity", position=position_dodge())

####### TOP 3 MOST WINNING TEAMS EACH YEAR
mostwinningteam <- data%>%group_by(YEAR,TEAM)%>%
  summarize(total = sum(W))%>%
  arrange(YEAR,-total)%>%
  slice(1:3)

mostwinningteam

ggplot(data=mostwinningteam, aes(x= YEAR, y=total, fill=TEAM)) +
  geom_bar(stat="identity", position=position_dodge())

######## MOST EFFICIENT TEAMS BASED ON TOURNAMENT APPRERANCES

mostefficient <- data%>%group_by(TEAM)%>%
  summarize(total = sum(MADETOURN))%>%
  arrange(-total)%>%
  slice(1:7)

mostefficient

usemosteff <- c(mostefficient$TEAM)

####### MOST EFFECIENT TEAMS PERFORMANCE 

trendsinbest <- data%>%
  filter(TEAM %in% usemosteff)%>%
  group_by(YEAR)%>%
  summarize(avgadjoe = mean(ADJOE),avgadjde = mean(ADJDE))


trendsinbest

ggplot(data= trendsinbest) + geom_smooth(aes(x=YEAR, y=ADJOE))

ggplot(data = trendsinbest, aes(x=YEAR,y=ADJOE,color=TEAM, group=TEAM)) +
  geom_line()

ggplot(data = trendsinbest, aes(x=YEAR,y=ADJDE,color=TEAM, group=TEAM)) +
  geom_line()

ggplot(trendsinbest, aes(x=YEAR)) + 
  geom_line(aes(y = avgadjoe),color = 'red') + 
  geom_line(aes(y = avgadjde),color = 'blue') 

### PCA TEST

pcatest = data[,!(colnames(data)%in% c("WINSEAS","MADETOURN","aboveavg"))]

names(pcatest)

res.pca <- prcomp(pcatest, scale = TRUE)

summary(res.pca)

fviz_eig(res.pca)

fviz_pca_var(res.pca,col.var = "contrib", gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),repel=TRUE)

### PREDICTOR AND DIMENSION REDUCTION

data = data[,!(colnames(data)%in% c("TEAM","CONF","POSTSEASON","SEED"))]

colnames(data)[colnames(data) %in% c("2P_O", "3P_O","2P_D","3P_D")] <- c("TWOPT_O", "THREEPT_O","TWOPT_D","THREEPT_D")

set.seed(1)

train=sample(c(TRUE,FALSE),prob=c(0.6,0.4),nrow(data),replace=TRUE)

test=!train

train.df=data[train,] 

test.df=data[-train,]


model1=regsubsets(W~.,data[train,],nvmax=20,method="exhaustive")
test.mat=model.matrix(W~.,data[test,])

val.errors=numeric(20)
for(i in 1:20){
  coefi=coef(model1,id=i)
  pred=test.mat[,names(coefi)]%*%coefi
  val.errors[i]=mean((data$W[test]-pred)^2)
}
val.errors
best=which.min(val.errors)
best
coef(model1,best)

trainmodel1 <-lm(formula = W ~ ADJOE + ADJDE + BARTHAG + EFG_O + EFG_D + TOR + TORD + ORB + DRB + FTR + FTRD + TWOPT_O + THREEPT_O + THREEPT_D + ADJ_T + WAB + YEAR + WINSEAS + aboveavg, data = train.df)

trainvstest<- predict(trainmodel1,test.df)
accuracy(trainvstest,test.df$W)

####### LOGISTIC REGRESSION TO PREDICT WINSEASON BASED ON OFFENSIVE DATA

model2 <- glm(formula = WINSEAS ~ YEAR + EFG_O + TOR + ORB + FTR + TWOPT_O + THREEPT_O, data = data, family = binomial)

summary(model2)

ggplot(data= model2) + geom_smooth(aes(x=YEAR, y=WINSEAS), method="glm",method.args = list(family = "binomial"))

ggplot(data=model2, aes(x=YEAR, y=WINSEAS)) +
  geom_smooth(method="glm", method.args = list(family = "binomial")) +
  stat_summary(data=model2, aes(x=round(YEAR,0), y=WINSEAS), fun=mean,
               geom="point") +
  scale_x_continuous(breaks=seq(2013, 2019, by=1), minor_breaks=NULL) +
  scale_y_continuous(labels = scales::percent) +
  theme_bw()

####### INTERPRETATIONS OF LOGISTIC REGRESSION

# At first glance we see two things that stand out. TOR has is negatively correlated with WINSEAS which is a good sign because we would expect the higher the turnover rate, the worse a team would do. Oddly we see that THREEPT_O is also negatively correlated, which seems odd that the more three points would have a negative effect on the odds of a winning season. 

LOG.ODDS <- predict(model2)
ODDS <- exp(LOG.ODDS)
PredictedProbability <- ODDS / (1 + ODDS)

PredictedProbability

MyROC <- roc(data$WINSEAS, PredictedProbability)

plot(MyROC)

coords(MyROC,"best")

MyROC

####### LINEAR REGRESSION TO FIND RELATIONS OF OFFENSIVE DATA

model3 <- lm(formula = W ~ YEAR + EFG_O + TOR + ORB + FTR + TWOPT_O + THREEPT_O, data = data)

summary(model3)

predict(model3, data, interval="prediction", level = .95)

####### LINEAR REGRESSION TO FIND RELATION OF DEFENSIVE DATA

model4 <- lm(formula = W ~ YEAR + EFG_D + TORD + DRB + FTRD + TWOPT_D + THREEPT_D, data = data)

summary(model4)

predict(model4, data, interval="prediction", level = .95)

####### COMPARING OFFENSIVE VS DEFENSIVE MODELS

summary(model3)

summary(model4)

####### INTERPRETATIONS OF COMPARISON

# When comparing both offensive and defensive linear models, given the higher r-squared of the offensive model, we can conclude that this model fits the data better. 


```

### Top 10 Teams by Total Wins
```{r TopTenTeams}
TopTeams <- ggplot(TeamWins, aes(x = reorder(TEAM, -TotalWins), y = TotalWins)) +
  geom_bar(aes(fill =(Conference)), stat = 'identity') +
  scale_fill_brewer(palette = 'Set2', name = 'Conference') +
  xlab('Teams') +
  ylab('Wins 2013-2019') +
  ggtitle('Top 10 Teams') +
  theme(axis.text.x  = element_text(angle = 65, vjust = .55))

plot(TopTeams)

```

### Top 10 Conferences by Total Wins
```{r Top10Conferences}
TopConference <- ggplot(TopWins, aes(x = reorder(CONF, -TotalWins), y = TotalWins)) +
  geom_bar(aes(fill =(CONF)), stat = 'identity') +
  xlab('Conference') +
  ylab('Wins 2013-2019') +
  ggtitle('Top 10 Conferences') +
  theme(axis.text.x  = element_text(angle = 65, vjust = .55))

plot(TopConference)

```


### Boxplot of free throwrates by conference
```{r ftrboxplot}
boxplot(ftrrate~CONF,data=conferenceclean, main="Free Throw Rate (How often the given team shoots Free Throws)", xlab="Conference", ylab="ftr rate")
```




## Modeling

Linear and logistic models


### Linear Model that predicts win based on a teams performance relative to their conference
```{r conferencewin}
confmodwinrate <- lm(W~efgrate+ftrrate+orbrate+efgdrate+torrate+tordrate+drbrate+ftrdrate,conferenceclean)
summary(confmodwinrate)



```


### Binomial model predicting whether team makes it to the final based on their conference avg and team's relative position
```{r conferencefinal}

isfinal <- glm(finals~ftrdrate+orbrate+avgefg+avgefgd+avgtord+as.factor(YEAR), data=conferenceclean, family="binomial")
summary(isfinal)

```


## Predictions

Put predictions using the logistic models here

### Prediction of team final by conference model
```{r predictions}
LOfinal <- predict(isfinal) 
ODDSfinal <- exp(LOfinal)
PredProbFinal <- ODDSfinal / (1 + ODDSfinal)
FinalsLogPred <- ifelse(PredProbFinal > 0.007562012, "Finalist", "Not Finalist")
table(FinalsLogPred, conferenceclean$finals)

```


## Conclusions

Linear: 1 unit increase
basic run down (significant, direction, etc)

Logistic:
Accuracy (ROC curve)


### Binomial model predicting whether team will make it finals based on conference performance

```{r conference final accuracy}
#accuracy of team final by conference model
FinalROC <- roc(conferenceclean$finals, PredProbFinal)
plot(FinalROC)
```

This model has a very high AUC

This model has a high sensitivity (~0.95)--accuracy at correctly predicting finalists. The model has a fairly high specificity (~0.8)--accuracy at estimating teams that don't make it to the final four. The model has a fairly high accuracy overall (~0.8)   

### Conference win linear model interpretation

Coefficients (all are ratio of team’s performance to their conference performance)

•	Positive coefficients – all offensive rates, when the team does well against the opposing team 

+	efgrate (Effective Field Goal Percentage Shot) 

+	ftrrate (How often the given team shoots Free Throws)

+	orbrate (Offensive Rebound Rate)

+	tordrate (Steal Rate)

•	Negative coefficients – all defensive rates, when the opposing team performs well against the examined

+	efgdrate (Effective Field Goal Percentage Allowed) 

+	ftrdrate (Free Throw Rate Allowed)

+	drbrate (Offensive Rebound Rate Allowed)

+	torrate (Turnover Rate)

•	One interpretation: For teams with the same ratio of team performance to conference performance for effective field goal shot & allowed, offensive rebound rate and rate allowed, free throw rate and rate allowed and turnover rate, an increase in the steal rate of 0.1 (performing an additional 10% better than your conference) is associated with ~1.5 additional wins

•	R^2^ is 0.69


## Predicting success in the NCAA Tournament based on Adjusted Offensive Efficiency 

```{r}
# Group Data by Year and Rank Teams Based on ADJOE
data <- CBB_ALL
new_data <- data %>%
  arrange(YEAR, ADJOE) %>%
  group_by(YEAR) %>%
  mutate(RNK = rank(-ADJOE, ties.method = 'first'))
View(new_data)

# Create Dummy Variable to Determine if Team Reached Final Four
new_data <- new_data %>%
  mutate(FINISH = ifelse(POSTSEASON %in% c('Champions', '2ND', 'F4'), 1, 0))
View(new_data)

# Create DataFrame that Only Includes Teams that Reached Final Four
f4 <- new_data %>%
  filter(FINISH == 1)
View(f4)

# View the Average ADJOE Rank of Teams That Reached Final Four by Year
av_rank <- f4 %>%
  group_by(YEAR) %>%
  summarise_at(vars(RNK), list(name = mean))
av_rank

attach(f4)
plot(RNK, ADJOE, pch = 19)
abline(lm(ADJOE ~ RNK), col = 'red')
lines(lowess(RNK, ADJOE), col = 'blue')

# Plot Comparing ADJOE and Rank of ADJOE of Teams in Final Four
ggplot(f4) +
 geom_point(aes(RNK, ADJOE, colour = as.character(YEAR))) +
  geom_vline(xintercept = 20, linetype = 'dashed', color = 'red') +
  ggtitle("Final Four Teams", subtitle = '(2013 - 2019)') +
  xlab('ADJOE Rank') + 
  ylab('Adj. Offensive Efficiency') +
  labs(color = "YEAR")
```

### Predicting if Teams will Reach the Final Four Based On ADJOE

```{r}
model <- glm(FINISH ~ ADJOE, family = binomial, data = new_data)
summary(model)

prob_of_y <- exp(0.30380)
prob_of_y
```

### Interpretation from Model

• Predicted Probability of Reaching the Final Four: As ADJOE increases by 1, the Predicted Probability of Reaching the Final Four increases by 1.35 times. 

• Example: Team A has an ADJOE of 120 and Team B has an ADJOE of 125. 
  - Predicted Probability of Team A reaching the Final Four = 14.35%
  - Predicted Probability of Team B reaching the Final Four = 43.36%
  
### Conclusion

• When filling out your bracket next year, if the teams that you choose to reach the Final Four do not rank in the top 20 for ADJOE, CREATE A NEW BRACKET!!!